---
title: "A"
output: html_document
---

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology, network analysis and gene set enrichment analysis using ClusterProfiler.

Update: 2nd August 2019

Analysis - simplified BP only for Felce only, Zhao only and Both (up/downregulated)

Note: used simplify() for every analysis - if editing and want to Knit to HTML, will take a long time!


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)
library(fgsea)
library(readr)

#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
# Data editing to create separate gene lists
```{r}

#Downregulated genes
A_down <- read_csv("total_results/A_down.csv")

#Create Felce only - remove anything where Zhao_LFC does not equal 0. (The ! at the front is for removing things!)
A_down_felce <- A_down[!(A_down$zhao_LFC !=0),]

#Create Zhao only - remove anything where Felce_LFC does not equal 0
A_down_zhao <- A_down[!(A_down$felce_LFC !=0),]

#Create Both - include anything where both LFC are smaller than 0
A_down_both <- A_down[(A_down$felce_LFC < 0 & A_down$zhao_LFC < 0),]

write.csv(A_down_felce,"results/A_down_felce.csv")
write.csv(A_down_zhao,"results/A_down_zhao.csv")
write.csv(A_down_both,"results/A_down_both.csv")

#Upregulated genes
A_up <- read_csv("total_results/A_up.csv")

#Create Felce only - remove anything where Zhao_LFC does not equal 0. (The ! at the front is for removing things!)
A_up_felce <- A_up[!(A_up$zhao_LFC !=0),]

#Create Zhao only - remove anything where Felce_LFC does not equal 0
A_up_zhao <- A_up[!(A_up$felce_LFC !=0),]

#Create Both - include anything where both LFC are larger than 0
A_up_both <- A_up[(A_up$felce_LFC > 0 & A_up$zhao_LFC > 0),]

write.csv(A_up_felce,"results/A_up_felce.csv")
write.csv(A_up_zhao,"results/A_up_zhao.csv")
write.csv(A_up_both,"results/A_up_both.csv")



```

# Gene Ontology

## Felce only

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
A_up_felce$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(A_up_felce$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
A_up_felce <- A_up_felce[is.na(A_up_felce$Entrez.Gene)==FALSE,]

#Remove duplicated entries
A_up_felce <- A_up_felce[!duplicated(A_up_felce$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Felce Upregulated genes, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(A_up_felce$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(ego)

#Only 6 terms - don't need to simplify!
#ego <- simplify(ego)

#Export list of biological processes and genes involved
A_up_felce_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(A_up_felce_BP, file="GO/A_up_felce_BP_simplified.csv")

```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```

```{r, fig.height=6, fig.width=8}
emapplot(ego)

```

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- A_up_felce$felce_LFC
#set name of object
names(logFC) <- A_up_felce$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 10, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```


## Zhao only

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
A_down_zhao$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(A_down_zhao$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
A_down_zhao <- A_down_zhao[is.na(A_down_zhao$Entrez.Gene)==FALSE,]

#Remove duplicated entries
A_down_zhao <- A_down_zhao[!duplicated(A_down_zhao$Entrez.Gene),]


#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
A_up_zhao$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(A_up_zhao$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
A_up_zhao <- A_up_zhao[is.na(A_up_zhao$Entrez.Gene)==FALSE,]

#Remove duplicated entries
A_up_zhao <- A_up_zhao[!duplicated(A_up_zhao$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Zhao Downregulated, Biological Process
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(A_down_zhao$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(ego)

#Only 2 terms, don't need to simplify!
#ego <- simplify(ego)


#Export list of biological processes and genes involved
A_down_zhao_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(A_down_zhao_BP, file="GO/A_down_zhao_BP_simplified.csv")


```

```{r,fig.height=4, fig.width=6}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)


```


```{r, fig.height=8, fig.width=10}
emapplot(ego)

```

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- A_down_zhao$zhao_LFC
#set name of object
names(logFC) <- A_down_zhao$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 10, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

### Gene Ontology (GO) over-representation test - Zhao Upregulated, Biological Process
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(A_up_zhao$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

ego <- simplify (ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
A_up_zhao_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(A_up_zhao_BP, file="GO/A_up_zhao_BP_simplified.csv")


```


```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```


```{r, fig.height=12, fig.width=14}
emapplot(ego)

```


```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- A_up_zhao$zhao_LFC
#set name of object
names(logFC) <- A_up_zhao$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 10, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```


## Both

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
A_up_both$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(A_up_both$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
A_up_both <- A_up_both[is.na(A_up_both$Entrez.Gene)==FALSE,]

#Remove duplicated entries
A_up_both <- A_up_both[!duplicated(A_up_both$Entrez.Gene),]

```


### Gene Ontology (GO) over-representation test - Both Upregulated, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(A_up_both$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
A_up_both_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(A_up_both_BP, file="GO/A_up_both_BP_simplified.csv")


```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```


```{r, fig.height=8, fig.width=10}
emapplot(ego)

```


```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- A_up_both$felce_LFC
#set name of object
names(logFC) <- A_up_both$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 10, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

