---
title: "D"
output: html_document
---

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology, network analysis and gene set enrichment analysis using ClusterProfiler.

NEED TO UPDATE FOR COMPARISON D

Analysis - simplified BP only for Lucic only, Zhao only and Both (up/downregulated). Check original ego object first to see how many GO terms (dim(ego)), if more than 100 then use simplify function.


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)
library(fgsea)
library(readr)

#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
# Data editing to create separate gene lists
```{r}

#Downregulated genes
D_down <- read_csv("total_results/D_down.csv")

#Create Lucic only - remove anything where Zhao_LFC does not equal 0. (The ! at the front is for removing things!)
D_down_lucic <- D_down[!(D_down$zhao_LFC !=0),]

#Create Zhao only - remove anything where Lucic_LFC does not equal 0
D_down_zhao <- D_down[!(D_down$lucic_LFC !=0),]

#Create Both - include anything where both LFC are smaller than 0
D_down_both <- D_down[(D_down$lucic_LFC < 0 & D_down$zhao_LFC < 0),]

write.csv(D_down_lucic,"results/D_down_lucic.csv")
write.csv(D_down_zhao,"results/D_down_zhao.csv")
write.csv(D_down_both,"results/D_down_both.csv")

#Upregulated genes
D_up <- read_csv("total_results/D_up.csv")

#Create Lucic only - remove anything where Zhao_LFC does not equal 0. (The ! at the front is for removing things!)
D_up_lucic <- D_up[!(D_up$zhao_LFC !=0),]

#Create Zhao only - remove anything where Lucic_LFC does not equal 0
D_up_zhao <- D_up[!(D_up$lucic_LFC !=0),]

#Create Both - include anything where both LFC are larger than 0
D_up_both <- D_up[(D_up$lucic_LFC > 0 & D_up$zhao_LFC > 0),]

write.csv(D_up_lucic,"results/D_up_lucic.csv")
write.csv(D_up_zhao,"results/D_up_zhao.csv")
write.csv(D_up_both,"results/D_up_both.csv")

```

# Gene Ontology

## Lucic only

Filter data
```{r}
#Downpregulated genes
#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_down_lucic$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_down_lucic$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_down_lucic <- D_down_lucic[is.na(D_down_lucic$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_down_lucic <- D_down_lucic[!duplicated(D_down_lucic$Entrez.Gene),]


#Upregulated genes
#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_up_lucic$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_up_lucic$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_up_lucic <- D_up_lucic[is.na(D_up_lucic$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_up_lucic <- D_up_lucic[!duplicated(D_up_lucic$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Lucic Downregulated genes, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_down_lucic$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID
dim(ego)

#ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_down_lucic_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_down_lucic_BP, file="GO/D_down_lucic_BP.csv")

```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```

```{r, fig.height=6, fig.width=8}
emapplot(ego)

```

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_down_lucic$lucic_LFC
#set name of object
names(logFC) <- D_down_lucic$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

### Gene Ontology (GO) over-representation test - Lucic Upregulated genes, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_up_lucic$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID
dim(ego)

ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_up_lucic_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_up_lucic_BP, file="GO/D_up_lucic_BP_simplified.csv")

```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```

```{r, fig.height=6, fig.width=8}
emapplot(ego)

```

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_up_lucic$lucic_LFC
#set name of object
names(logFC) <- D_up_lucic$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

## Zhao only

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_down_zhao$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_down_zhao$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_down_zhao <- D_down_zhao[is.na(D_down_zhao$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_down_zhao <- D_down_zhao[!duplicated(D_down_zhao$Entrez.Gene),]


#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_up_zhao$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_up_zhao$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_up_zhao <- D_up_zhao[is.na(D_up_zhao$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_up_zhao <- D_up_zhao[!duplicated(D_up_zhao$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Zhao Downregulated, Biological Process
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_down_zhao$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID
dim(ego)

ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_down_zhao_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_down_zhao_BP, file="GO/D_down_zhao_BP_simplified.csv")


```

```{r,fig.height=4, fig.width=6}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)


```


```{r, fig.height=8, fig.width=10}
emapplot(ego)

```

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_down_zhao$zhao_LFC
#set name of object
names(logFC) <- D_down_zhao$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

### Gene Ontology (GO) over-representation test - Zhao Upregulated, Biological Process
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_up_zhao$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

dim(ego)

#ego <- simplify (ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_up_zhao_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_up_zhao_BP, file="GO/D_up_zhao_BP.csv")


```


```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```


```{r, fig.height=12, fig.width=14}
emapplot(ego)

```


```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_up_zhao$zhao_LFC
#set name of object
names(logFC) <- D_up_zhao$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```


## Both

Filter data
```{r}
#Downregulated
#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_down_both$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_down_both$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_down_both <- D_down_both[is.na(D_down_both$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_down_both <- D_down_both[!duplicated(D_down_both$Entrez.Gene),]


#Upregulated
#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
D_up_both$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(D_up_both$gene), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
D_up_both <- D_up_both[is.na(D_up_both$Entrez.Gene)==FALSE,]

#Remove duplicated entries
D_up_both <- D_up_both[!duplicated(D_up_both$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Both Downregulated, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_down_both$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

dim(ego)

#ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_down_both_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_down_both_BP, file="GO/D_down_both_BP.csv")


```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```


```{r, fig.height=8, fig.width=10}
emapplot(ego)

```


```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_down_both$lucic_LFC
#set name of object
names(logFC) <- D_down_both$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

### Gene Ontology (GO) over-representation test - Both Upregulated, Biological Process - CONTINUE FROM HERE 7TH AUGUST

```{r}
#Change Entrez IDs from numbers to characters

geneset <- as.character(D_up_both$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

dim(ego)

ego <- simplify(ego)

#dimensions - number of GO terms
dim(ego)

#Export list of biological processes and genes involved
D_up_both_BP <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(D_up_both_BP, file="GO/D_up_both_BP_simplified.csv")


```

```{r,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

```


```{r, fig.height=8, fig.width=10}
emapplot(ego)

```


```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- D_up_both$lucic_LFC
#set name of object
names(logFC) <- D_up_both$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```


