---
title: "A"
output: html_document
---

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology, network analysis and gene set enrichment analysis using ClusterProfiler.

Update: 3rd December 2019

Analysis - 2 gene lists, Jurkats only and CD4+ cells only

Note: used simplify() for analysis - if editing and want to Knit to HTML, will take a long time!


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)
library(fgsea)
library(readr)

#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
# Data editing to read in separate gene lists
```{r}

jurkat <- read.csv("total_results/jurkats_only.csv", header=FALSE)
names(jurkat)[1] <- "ensembl"

cd4 <- read.csv("total_results/cd4_only.csv", header=FALSE)
names(cd4)[1] <- "ensembl"


```

# Gene Ontology

## Jurkat only

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
jurkat$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(jurkat$ensembl), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
jurkat <- jurkat[is.na(jurkat$Entrez.Gene)==FALSE,]

#Remove duplicated entries
jurkat <- jurkat[!duplicated(jurkat$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - Jurkat only, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

jurkat_geneset <- as.character(jurkat$Entrez.Gene)

#This will show Entrez IDs only
head(jurkat_geneset)

#This will take a little while to run
jurkat_ego <- enrichGO(gene = jurkat_geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(jurkat_ego)

#simplify
simp_jurkat_ego <- simplify(jurkat_ego)

dim(simp_jurkat_ego)

#Export list of biological processes and genes involved
jurkat_BP <- data.frame(simp_jurkat_ego$ID, simp_jurkat_ego$Description, simp_jurkat_ego$p.adjust, simp_jurkat_ego$geneID)

write.csv(jurkat_BP, file="GO/jurkat_BP_simplified.csv")

```

```{r,fig.height=6, fig.width=8}
barplot (simp_jurkat_ego, showCategory = 10)

dotplot (simp_jurkat_ego, showCategory=10)

```

```{r, fig.height=10, fig.width=10}
emapplot(simp_jurkat_ego)

```

```{r, fig.height=12, fig.width=14}
#no logfc information available

cnetplot(simp_jurkat_ego, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```


## CD4 only

Filter data
```{r}

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
cd4$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(cd4$ensembl), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
cd4 <- cd4[is.na(cd4$Entrez.Gene)==FALSE,]

#Remove duplicated entries
cd4 <- cd4[!duplicated(cd4$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test - CD4 only, Biological Process

```{r}
#Change Entrez IDs from numbers to characters

cd4_geneset <- as.character(cd4$Entrez.Gene)

#This will show Entrez IDs only
head(cd4_geneset)

#This will take a little while to run
cd4_ego <- enrichGO(gene = cd4_geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(cd4_ego)

#98 terms - no need to simplify
#simp_cd4_ego <- simplify(cd4_ego)


#Export list of biological processes and genes involved
cd4_BP <- data.frame(cd4_ego$ID, cd4_ego$Description, cd4_ego$p.adjust, cd4_ego$geneID)

write.csv(cd4_BP, file="GO/cd4_BP.csv")

```

```{r,fig.height=6, fig.width=8}
barplot (cd4_ego, showCategory = 10)

dotplot (cd4_ego, showCategory=10)

```

```{r, fig.height=10, fig.width=10}
emapplot(cd4_ego)

```

```{r, fig.height=12, fig.width=12}
#no logfc information available

cnetplot(cd4_ego, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated

```

