---
title: "GO"
output: html_document
---
Update: 3rd July 2019

# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology, network analysis and gene set enrichment analysis using FGSEA (https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html) - doesn't produce as nice enrichment plots as GSEA GUI, also have to do individually for each DE gene list.


```{r, include=FALSE}

library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pathview)
library(fgsea)

#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.

```
## Resting vs Activated CD4 T Cells - using DESeq2 gene list

### Filter data (process data)
The data contains double entries or NA entries. We need to remove these for downstream analysis.

```{r 2h import data}

#Import significant results from DESeq2 analysis 

library(readr)
sig <- read_csv("results/rest_vs_activated_sig.csv")

#Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are

sig$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(sig$Row.names), keytype="ENSEMBL", column="ENTREZID")

#Keep only rows from table without NAs
sig <- sig[is.na(sig$Entrez.Gene)==FALSE,]

#Remove duplicated entries
sig <- sig[!duplicated(sig$Entrez.Gene),]

```

### Gene Ontology (GO) over-representation test
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r GO test}
#Change Entrez IDs from numbers to characters

geneset <- as.character(sig$Entrez.Gene)

#This will show Entrez IDs only
head(geneset)

#This will take a little while to run
ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID

#dimensions - number of GO terms
dim(ego)

head(ego)
#GeneRatio - how may times appear in gene list
#BgRatio - how many times appear in universe

#Export list of biological processes and genes involved
GOgenes <- data.frame(ego$ID, ego$Description, ego$p.adjust, ego$geneID)

write.csv(GOgenes, file="GO/rest_vs_activated_BP.csv")


```
### Showing the ego object in graphical form

```{r barplot and dotplot,fig.height=6, fig.width=8}
barplot (ego, showCategory = 10)

dotplot (ego, showCategory=10)

#can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))

```


We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r GO terms relationship, fig.height=12, fig.width=14}
emapplot(ego)

```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r genes in GO terms, fig.height=20, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC <- sig$log2FoldChange
#set name of object
names(logFC) <- sig$Entrez.Gene
head(logFC)

cnetplot(ego, foldChange = logFC, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated


```

### Gene Set Enrichment Analysis (GSEA)
GSEA is a method for testing the expression of focused gene sets in gene expression data. In contrast to the over-representation tests we have seen, GSEA considers all genes in the data, so there is no need for an arbitary cutoff.

```{r reactome barplot}
#For running GSEA with the fgsea() function, we need to specify a list of gene sets (could also be pathways) and a metric that will be used to rank the data. Here we are using fold change and reactome pathways.

pathways <- reactomePathways (names(logFC))

#"pathways" has to be named list of gene sets
length(pathways)

#Here we are using logFC for ranking the genes
ranks <- logFC

#This is to demonstrate ranked data - ranked gene list
barplot(sort(ranks, decreasing=T))


```
### Exploring the results - Reactome Pathways

``` {r reactome plots}
#fgsea will perform GSEA
#install packages stats, DOSE, colorspace

gsea_results <- fgsea(pathways = pathways,
                      stats = ranks,
                      minSize =10,
                      maxSize = 500,
                      nperm=10000)

colnames(gsea_results)

dim(gsea_results)

head(gsea_results)

#results from fgsea function are in same order as pathways. We might need to filter or rank them depending on ES and padj. 

significant <-gsea_results$padj <=0.05

gsea_significant<-gsea_results[significant,]


#order results based on ES score
ES_order <- order(gsea_significant$ES, decreasing=TRUE)

gsea_ordered <- gsea_significant[ES_order]

gsea_ordered

#to export, copy and paste table into Excel - doesn't like write.csv command as list of genes in last column.

#visualise GSEA results for some of these pathways - top 5

plotEnrichment(pathways[["Unwinding of DNA"]], ranks)
plotEnrichment(pathways[["Response to metal ions"]], ranks)
plotEnrichment(pathways[["Polo-like kinase mediated events"]], ranks)
plotEnrichment(pathways[["Chk1/Chk2(Cds1) mediated inactivation of Cyclin B:Cdk1 complex"]], ranks)
plotEnrichment(pathways[["RNA Polymerase I Promoter Opening"]], ranks)


```

### Using genesets from MsigDB for GSEA

The MSigDB is a collection of annotated gene sets. Different parts of MSigDB are available in  [http://software.broadinstitute.org/gsea/downloads.jsp](http://software.broadinstitute.org/gsea/downloads.jsp).


```{r load hallmark}
#downloaded hallmark gene sets, Entrez IDs

h_all <- read.gmt(gmtfile="MSigDB/h.all.v6.2.entrez.gmt")


dim(h_all)

head(h_all)

#create an empty list first
pathways<-list()

#separate gene sets and add as named elements of the list
for (gs in unique (h_all$ont)){
  gs_ids <- h_all$ont == gs
  pathways[[gs]] <- h_all$gene[gs_ids]
}

```


Now, we can do GSEA with the pathways from MSigDB.

```{r hallmark plots}

gsea_results <- fgsea(pathways=pathways,
                      stats=ranks,
                      minSize=10,
                      maxSize=500,
                      nperm=10000)
                     

dim(gsea_results)

#keep significant genesets and test order based on ES

significant <- gsea_results$padj <0.05

gsea_significant <- gsea_results[significant,]

ES_order<- order(gsea_significant$ES, decreasing=TRUE)

gsea_ordered<-gsea_significant[ES_order,]

gsea_ordered

#Plot GSEA plots - top 5
plotEnrichment(pathways[["HALLMARK_E2F_TARGETS"]],ranks)
plotEnrichment(pathways[["HALLMARK_G2M_CHECKPOINT"]],ranks)
plotEnrichment(pathways[["HALLMARK_MYC_TARGETS_V2"]],ranks)
plotEnrichment(pathways[["HALLMARK_MYC_TARGETS_V1"]],ranks)
plotEnrichment(pathways[["HALLMARK_MITOTIC_SPINDLE"]],ranks)

#copy and paste gsea_ordered into Excel spreadsheet

#If want to find out genes involved - copy Entrez IDs from Excel
#entrez_ids <- c("1282","1284","4323","131578","1809","7058","7474","2335","2619","627","7857","2191","1009","3371","2200","4312","4638","6695","5396","23705","11167","1303","3569")

#Convert to symbols using Hs database, then copy and paste into Excel
#symbols <- mapIds(org.Hs.eg.db, keys=as.character(entrez_ids), keytype="ENTREZID", column="SYMBOL")

#symbols

```

