---
title: "Comparison C"
output: html_document
---


Update: 8th July 2019

# Comparison C: Felce resting v stimulated VS Lucic resting v activated +/- 1 log2 fold change
Using DESeq2 total gene lists

```{r, include=FALSE}
source('functions.R')
library(org.Hs.eg.db)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(limma)
library(pheatmap)
```

```{r upload DESeq2 genes}

felce <- read_csv("C_rest_vs_stim_res.csv")
lucic <- read_csv ("C_rest_vs_activated_res.csv")

#Use Row.names to be actual rownames (Ensembl IDs)

felce_names <- felce$Row.names
lucic_names <- lucic$Row.names

rownames(felce) <- make.names (felce_names, unique=TRUE)
rownames(lucic) <- make.names(lucic_names, unique=TRUE)


```

```{r separate up or downregulated}

#downregulated

felce_down <- felce %>%
  rownames_to_column("gene") %>%
  filter(log2FoldChange < -1)

lucic_down <- lucic %>%
  rownames_to_column("gene") %>%
  filter(log2FoldChange < -1)

#upregulated

felce_up <- felce %>%
  rownames_to_column("gene") %>%
  filter (log2FoldChange > 1)

lucic_up <- lucic %>%
  rownames_to_column("gene") %>%
  filter (log2FoldChange > 1)


```


# Venn diagrams
### Shared downregulated genes 

```{r venn down}
library(reshape)
library(reshape2)


felce_down <- felce_down %>% 
  select(gene, log2FoldChange)
lucic_down <- lucic_down %>% 
  select(gene, log2FoldChange)

merge_down <- merge(felce_down, lucic_down, by="gene", all = TRUE)

merge_down <- merge_down %>% 
  mutate(felce_LFC = log2FoldChange.x, lucic_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_down[is.na(merge_down)] <- 0

felce <- (merge_down$felce_LFC < -1)
lucic <- (merge_down$lucic_LFC < -1)

c2 <- cbind(felce, lucic)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn/C_downregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn/C_downregulated_venn.tiff")



# Next subset the overlapping genes for all

merge_down <- ensembl_to_symbol(merge_down, "gene")
down <- merge_down[complete.cases(merge_down),]


write.csv(down, file="venn/C_down.csv")
```

### Shared upregulated genes

```{r venn up}

library(reshape)
library(reshape2)


felce_up <- felce_up %>% 
  select(gene, log2FoldChange)
lucic_up <- lucic_up %>% 
  select(gene, log2FoldChange)

merge_up <- merge(felce_up, lucic_up, by="gene", all = TRUE)

merge_up <- merge_up %>% 
  mutate(felce_LFC = log2FoldChange.x, lucic_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_up[is.na(merge_up)] <- 0

felce <- (merge_up$felce_LFC > 1)
lucic <- (merge_up$lucic_LFC > 1)

c2 <- cbind(felce, lucic)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn/C_upregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn/C_upregulated_venn.tiff")



# Next subset the overlapping genes for all

merge_up <- ensembl_to_symbol(merge_up, "gene")
up <- merge_up[complete.cases(merge_up),]


write.csv(up, file="venn/C_up.csv")

```

# Comparison C: Felce resting v stimulated VS Lucic resting v activated +/- 0 log2 fold change
Update: 18th July 2019

```{r upload genes}

felce <- read_csv("C_rest_vs_stim_res.csv")
lucic <- read_csv ("C_rest_vs_activated_res.csv")

#Use Row.names to be actual rownames (Ensembl IDs)

felce_names <- felce$Row.names
lucic_names <- lucic$Row.names

rownames(felce) <- make.names (felce_names, unique=TRUE)
rownames(lucic) <- make.names(lucic_names, unique=TRUE)


```

```{r separate up or down}

#downregulated

felce_down <- felce %>%
  rownames_to_column("gene") %>%
  filter(log2FoldChange < 0)

lucic_down <- lucic %>%
  rownames_to_column("gene") %>%
  filter(log2FoldChange < 0)

#upregulated

felce_up <- felce %>%
  rownames_to_column("gene") %>%
  filter (log2FoldChange > 0)

lucic_up <- lucic %>%
  rownames_to_column("gene") %>%
  filter (log2FoldChange > 0)


```


# Venn diagrams
### Shared downregulated genes 

```{r down}
library(reshape)
library(reshape2)


felce_down <- felce_down %>% 
  select(gene, log2FoldChange)
lucic_down <- lucic_down %>% 
  select(gene, log2FoldChange)

merge_down <- merge(felce_down, lucic_down, by="gene", all = TRUE)

merge_down <- merge_down %>% 
  mutate(felce_LFC = log2FoldChange.x, lucic_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_down[is.na(merge_down)] <- 0

felce <- (merge_down$felce_LFC < 0)
lucic <- (merge_down$lucic_LFC < 0)

c2 <- cbind(felce, lucic)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn/C_downregulated_venn_new.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn/C_downregulated_venn_new.tiff")



# Next subset the overlapping genes for all

merge_down <- ensembl_to_symbol(merge_down, "gene")
down <- merge_down[complete.cases(merge_down),]


write.csv(down, file="venn/C_down_new.csv")
```

### Shared upregulated genes 

```{r up}

library(reshape)
library(reshape2)


felce_up <- felce_up %>% 
  select(gene, log2FoldChange)
lucic_up <- lucic_up %>% 
  select(gene, log2FoldChange)

merge_up <- merge(felce_up, lucic_up, by="gene", all = TRUE)

merge_up <- merge_up %>% 
  mutate(felce_LFC = log2FoldChange.x, lucic_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_up[is.na(merge_up)] <- 0

felce <- (merge_up$felce_LFC > 0)
lucic <- (merge_up$lucic_LFC > 0)

c2 <- cbind(felce, lucic)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn/C_upregulated_venn_new.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn/C_upregulated_venn_new.tiff")



# Next subset the overlapping genes for all

merge_up <- ensembl_to_symbol(merge_up, "gene")
up <- merge_up[complete.cases(merge_up),]


write.csv(up, file="venn/C_up_new.csv")

```